Validation of an Actuator Line Model Coupled to a Dynamic Stall Model for Pitching Motions Characteristic to Vertical Axis Turbines

### 2016-03-02 paper.tex
New corrections
New corrections Goude - Mendoza

    Vertical Axis Wind Turbines (VAWTs) have been used for extracting of energy
    either from sea or wind. A simpler design, low cost of maintenance and the
    capability of working independent from the direction of the flow are some of
    the most important advantages over conventional Horizontal Axis Wind
    Turbines (HAWTs). Complex and unsteady fluid dynamics are not easy to model
    for simulation tools. One of the most relevant phenomena is the dynamic
    stall which is inherently to unsteady conditions, this work is focused in
    its study. Dynamic stall is usually used as a passive control for VAWT
    operating conditions, hence the importance of a good understanding. In this
    study, \textbf{a coupled model is implemented with the open-source CFD
    toolbox OpenFOAM for solving the governing equations; an actuator line model
    is used for solving velocity flow field and a dynamic stall model for force
    interactions.} Force coefficients obtained from the model are validated with
    experimental data of pithing airfoil in similar operating conditions as
    H-rotor type. Numerical results show reasonable agreement with experimental
    data for pitching motion. In general, simulations allow to characterize
    dynamic stall effects  and their magnitude.

@@

Operating conditions of VAWT are characterized as complex unsteady flows which
give a considerable challenge, both to describe using measurements and to
represent through simulation tools\cite{huyer1996unsteady}. Moreover, VAWT
blades are inherently exposed to cyclic variation in the angle of attack, which
gives cyclic blade forces and can give material fatigue damage. Accurate
modeling of the varying forces is therefore very important for the design of the
VAWT.

@@

\textbf{Actuator Line Model:} is a three-dimensional and unsteady aerodynamic
model developed by S{\o}rensen and Shen\cite{sorensen1999computation}, used to
study the flow around wind turbines. It is a combination of a solver of
Navier-Stokes equations with a so-called actuator line technique, in which
blades of the turbine are represented by a radial distribution of body forces
along lines. These forces are determined using a dynamic stall model commonly
based on empirical data.

\textbf{Dynamic Stall Modeling:} represented by the Leishman-Beddoes
model\cite{leishman1986generalised} with the modifications of Sheng et
al\cite{sheng2008modified} and Dyachuk\cite{dyachuk},
using the library turbinesFoam developed by Bachant\cite{bachant2015simulating}.
It is capable to calculate the unsteady lift,
pitching moment and drag, giving a physical description of the aerodynamics. It
have been validated with experimental data in\cite{leishman1989semi}.
The model is separated into three subsystems: an attached flow model for unsteady linear
airloads, a separated flow model for non-linear airloads and a dynamic stall
model for the airloads induced by the leading edge vortex.

A NACA0021 airfoil was tested at the Reynolds number of 1.000.000 (which is a
reasonable value for operating VAWT) during pithing motion similar to the motion
of the VAWT blade. Different amplitude degrees were investigated. In a VAWT with
fixed blades, the angle of attack is in function of the TSR of the turbine. A
periodic function is used to represent the variations of the angles of attack
which a VAWT blade experiences analog to the pitching motion,
\begin{eqnarray}
    \alpha = \arctan \left( \frac{\sin \theta}{\lambda + \cos \theta} \right)
\end{eqnarray}
Where $\theta$ is the azimuthal blade angle and $ \lambda $ represents the TSR,
\begin{eqnarray}
    \lambda =  \frac{\Omega R}{V}
\end{eqnarray}
The pitching blade experiments were carried out at Glasgow
University\cite{angell1988collected}\\

\begin{figure}[h]
\begin{minipage}{18pc}
\includegraphics[width=18pc]{Fig1.png}
\end{minipage}\hspace{2pc}%
\begin{minipage}{18pc}
\includegraphics[width=18pc]{Fig2.png}
\end{minipage}
\caption{\label{fig1}Normal (left) and tangential (right) force responses during pitching motions of NACA0021 airfoil with a maximum amplitude of 17.4\degree\ (analog to a $\lambda$ = 3.34).}
\end{figure}

\begin{figure}[h]
\begin{minipage}{18pc}
\includegraphics[width=18pc]{Fig3.png}
\end{minipage}\hspace{2pc}%
\begin{minipage}{18pc}
\includegraphics[width=18pc]{Fig4.png}
\end{minipage}
\caption{\label{fig2}Normal (left) and tangential (right) force responses during pitching motions of NACA0021 airfoil with a maximum amplitude of 22.6\degree\ (analog to a $\lambda$ = 2.60).}
\end{figure}

@@

For a TSR of 3.34 (Figure \ref{fig1}), the maximum magnitude of the angle of attack is at
17.4\degree\ which is in the stall region. Both CN and CT curves show the delay
of the flow reattachment, which is characteristic of the dynamic stall
phenomenon. Peaks of simulated values have close agreement with the measured
data.

@@

The coupled model was set-up for the operating conditions of a VAWT. The reliability of the results is acceptable, making a good agreement with
experimental data from the wind tunnel test, considering the complexity of the flows studied.

%It has been demonstrated how the coupled actuator line model can be coupled to
%a modified version of the Leishman-Beddoes model, and the results could be
%validated for similar operating conditions of a VAWT.

>>>>

    Vertical Axis Wind Turbines (VAWTs) have been used for extracting of energy
    either from sea or wind. A simpler design, low cost of maintenance and the
    capability of working independent from the direction of the flow are some of
    the most important advantages over conventional Horizontal Axis Wind
    Turbines (HAWTs). Complex and unsteady fluid dynamics are not easy to model
    for simulation tools. One of the most relevant phenomena is the dynamic
    stall which is inherently to unsteady conditions, this work is focused in
    its study. Dynamic stall is usually used as a passive control for VAWT
    operating conditions, hence the importance of a good understanding. In this
    study, \textbf{a coupled model is implemented with the open-source CFD
    toolbox OpenFOAM for solving the governing equations; an actuator line model
    is used for solving velocity flow field and a dynamic stall model for force
    interactions.} Force coefficients obtained from the model are validated with
    experimental data of pithing airfoil in similar operating conditions as
    H-rotor type. Numerical results show reasonable agreement with experimental
    data for pitching motion.

@@

Operating conditions of VAWTs are characterized as complex unsteady flows which
give a considerable challenge, both to describe using measurements and to
represent through simulation tools\cite{huyer1996unsteady}. Moreover, VAWT
blades are inherently exposed to cyclic variation in the angle of attack, which
gives cyclic blade forces and can give material fatigue damage. Accurate
modeling of the varying forces is therefore very important for the design of the
VAWT.

@@

\textbf{Actuator Line Model:} is a three-dimensional and unsteady aerodynamic
model developed by S{\o}rensen and Shen\cite{sorensen1999computation}, used to
study the flow around wind turbines. It is a combination of a solver of
Navier-Stokes equations with a so-called actuator line technique, in which
blades of the turbine are represented by a radial distribution of body forces
along lines. These forces are determined using a dynamic stall model commonly
based on empirical data. This work uses the library turbinesFoam developed by
Bachant\cite{bachant2015simulating}.

\textbf{Dynamic Stall Modeling:} is represented by the Leishman-Beddoes
model\cite{leishman1986generalised} with the modifications of Sheng et
al\cite{sheng2008modified} and Dyachuk\cite{dyachuk}.
It is capable to calculate the unsteady lift,
pitching moment and drag, giving a physical description of the aerodynamics. It
have been validated with experimental data in\cite{leishman1989semi}.
The model is separated into three subsystems: an attached flow model for unsteady linear
airloads, a separated flow model for non-linear airloads and a dynamic stall
model for the airloads induced by the leading edge vortex.

A NACA0021 airfoil was tested at the Reynolds number of 1.000.000 (which is a
reasonable value for operating VAWT) during pithing motion similar to the motion
of the VAWT blade. Different pitching amplitudes were investigated. In a VAWT with
fixed blades, the angle of attack is in function of the TSR of the turbine. A
periodic function is used to represent the variations of the angles of attack
which a VAWT blade experiences analog to the pitching motion,
\begin{eqnarray}
    \alpha = \arctan \left( \frac{\sin \theta}{\lambda + \cos \theta} \right),
\end{eqnarray}
where $\theta$ is the azimuthal blade angle and $ \lambda $ represents the TSR,
\begin{eqnarray}
    \lambda =  \frac{\Omega R}{V}.
\end{eqnarray}
The pitching blade experiments were carried out at Glasgow
University\cite{angell1988collected}.\\

\begin{figure}[h]
\begin{minipage}{18pc}
\includegraphics[width=18pc]{Fig1.png}
\end{minipage}\hspace{2pc}%
\begin{minipage}{18pc}
\includegraphics[width=18pc]{Fig2.png}
\end{minipage}
\caption{\label{fig1}Normal (left) and tangential (right) force coefficients during pitching motions of NACA0021 airfoil with a maximum amplitude of 17.4\degree\ (analog to $\lambda = 3.34$).}
\end{figure}

\begin{figure}[h]
\begin{minipage}{18pc}
\includegraphics[width=18pc]{Fig3.png}
\end{minipage}\hspace{2pc}%
\begin{minipage}{18pc}
\includegraphics[width=18pc]{Fig4.png}
\end{minipage}
\caption{\label{fig2}Normal (left) and tangential (right) force coefficients during pitching motions of NACA0021 airfoil with a maximum amplitude of 22.6\degree\ (analog to $\lambda = 2.60$).}
\end{figure}

@@

For a TSR of 3.34 (Figure \ref{fig1}), the maximum magnitude of the angle of attack is at
17.4\degree\ which is in the stall region. Both \textit{CN} and \textit{CT} curves show the delay
of the flow reattachment, which is characteristic of the dynamic stall
phenomenon. Peaks of simulated values have close agreement with the measured
data.

@@

In general, the model show good agreement with experimental data, considering the complexity
of the flows studied. This makes the model suitable for application in wind turbine
simulations.
